\documentclass[10pt]{article}
\usepackage{listings}
\usepackage{soul,ifthen}
\lstnewenvironment{code}[1][]{%
  \lstset{
    basicstyle=\ttfamily,
    columns=fullflexible,
    keepspaces=true,
    frame=left,
    escapechar=^,
    #1
  }%
}{}
\usepackage{fancyvrb}
\DefineVerbatimEnvironment{command}{Verbatim}{commandchars=\\\{\}}

\usepackage{amsthm}
\theoremstyle{remark}
\newtheorem{warning}{Warning}

\usepackage{xcolor} \newcommand{\jg}[1]{\textcolor{violet}{JG: #1}}
\newcommand{\mz}[1]{\textcolor{red}{MZ: #1}}


\newcommand{\taylorname}{{\bf taylor}}
\newcommand{\version}{2.0.0}

\newcommand{\myfloat}{{\tt MY\symbol{95}FLOAT}}
\newcommand{\myjet}{{\tt MY\symbol{95}JET}}
\newcommand{\myscal}{{\tt MY\symbol{95}SCAL}}

\newcommand{\inputfiles}[1]{%
\ifthenelse{\equal{#1}{sec:ex0}}{{\tt lorenz1.eq}}{%
\ifthenelse{\equal{#1}{sec:ex1}}{{\tt lorenz1.eq}}{%
\ifthenelse{\equal{#1}{sec:ex_params}}{{\tt perturbation.eq}}{%
\ifthenelse{\equal{#1}{sec:ex_lyap}}{{\tt lorenz2.eq}}{%
\ifthenelse{\equal{#1}{sec:ex_varieq}}{{\tt lorenz3.eq}}{%
\ifthenelse{\equal{#1}{sec:ex5}}{{\tt lorenz4.eq}}{%
\ifthenelse{\equal{#1}{sec:ex_omp}}{{\tt lorenz4.eq}}{%
\ifthenelse{\equal{#1}{sec:ex_equilibria}}{{\tt lorenz5.eq}}{%
\ifthenelse{\equal{#1}{sec:ex_myjetaccess}}{{\tt model.eq}}{%
\ifthenelse{\equal{#1}{sec:ex_driving}}{{\tt lorenz1.eq}}{%
ERROR!!!!!}}}}}}}}}}%
}
\newcommand{\odecfiles}[1]{%
\ifthenelse{\equal{#1}{sec:ex0}}{{\tt lorenz1.c}}{%
\ifthenelse{\equal{#1}{sec:ex1}}{{\tt lorenz1.c}}{%
\ifthenelse{\equal{#1}{sec:ex_params}}{{\tt perturbation.c}}{%
\ifthenelse{\equal{#1}{sec:ex_lyap}}{{\tt lorenz2.c}}{%
\ifthenelse{\equal{#1}{sec:ex_varieq}}{{\tt lorenz3.c}}{%
\ifthenelse{\equal{#1}{sec:ex5}}{{\tt lorenz4.c}}{%
\ifthenelse{\equal{#1}{sec:ex_omp}}{{\tt lorenz4.c}}{%
\ifthenelse{\equal{#1}{sec:ex_equilibria}}{{\tt lorenz5.c}}{%
\ifthenelse{\equal{#1}{sec:ex_myjetaccess}}{{\tt jdata.c}}{%
\ifthenelse{\equal{#1}{sec:ex_driving}}{{\tt lorenz.c}}{%
ERROR!!!!!}}}}}}}}}}%
}
\newcommand{\odehfiles}[1]{%
\ifthenelse{\equal{#1}{sec:ex0}}{{\tt taylor.h}}{%
\ifthenelse{\equal{#1}{sec:ex1}}{{\tt taylor.h}}{%
\ifthenelse{\equal{#1}{sec:ex_params}}{{\tt taylor.h}}{%
\ifthenelse{\equal{#1}{sec:ex_lyap}}{{\tt lorenz2.h}}{%
\ifthenelse{\equal{#1}{sec:ex_varieq}}{{\tt lorenz3.h}}{%
\ifthenelse{\equal{#1}{sec:ex5}}{{\tt lorenz4.h}}{%
\ifthenelse{\equal{#1}{sec:ex_omp}}{{\tt lorenz4.h}}{%
\ifthenelse{\equal{#1}{sec:ex_equilibria}}{{\tt lorenz5.h}}{%
\ifthenelse{\equal{#1}{sec:ex_myjetaccess}}{{\tt jdata.h}}{%
\ifthenelse{\equal{#1}{sec:ex_driving}}{{\tt taylor.h}}{%
ERROR!!!!!}}}}}}}}}}%
}
\newcommand{\mainfiles}[1]{%
\ifthenelse{\equal{#1}{sec:ex0}}{{\tt main\symbol{95}lrnz.c}}{%
\ifthenelse{\equal{#1}{sec:ex1}}{{\tt main\symbol{95}lrnz.c}}{%
\ifthenelse{\equal{#1}{sec:ex_params}}{{\tt main\symbol{95}params.c}}{%
\ifthenelse{\equal{#1}{sec:ex_lyap}}{{\tt main\symbol{95}lyap.c}}{%
\ifthenelse{\equal{#1}{sec:ex_varieq}}{{\tt main\symbol{95}varieq.c}}{%
\ifthenelse{\equal{#1}{sec:ex5}}{{\tt main4.c}}{%
\ifthenelse{\equal{#1}{sec:ex_omp}}{{\tt main\symbol{95}omp.c}}{%
\ifthenelse{\equal{#1}{sec:ex_equilibria}}{{\tt main\symbol{95}equilibria.c}}{%
\ifthenelse{\equal{#1}{sec:ex_myjetaccess}}{{\tt jdata\symbol{95}main.c}}{%
ERROR!!!!!}}}}}}}}}%
}

\newcommand{\inputfile}{}
\newcommand{\mainfile}{}
\newcommand{\odecfile}{}
\newcommand{\odehfile}{}

\setlength\parindent{0pt} % to remove indent

\title{\taylorname{} User's Manual\\[.5em] \normalsize version \version{}}

\author{Joan Gimeno {\tt <joan@maia.ub.es>} \and \`Angel Jorba {\tt
    <angel@maia.ub.es>} \and Maorong Zou {\tt <mzou@math.utexas.edu>}}
\date{September 9, 2022}

\voffset=-20mm       % vertical displacement
\oddsidemargin=3mm   % horizontal displacement (odd pages)
\evensidemargin=3mm  % horizontal displacement (even pages)
%
\textwidth=160mm
\textheight=225mm

\begin{document}
\maketitle

\vspace{-2em}\tableofcontents \newpage %to be removed?

\section{What is \taylorname{}} \label{sec:whatis}
\taylorname{} is an Ordinary Differential Equation (ODE) solver
generator.  It reads a system of ODEs and outputs an ANSI C routine
that performs a single step of numerical integration using the Taylor
method. Each step of integration chooses the step and the order
adaptively to keep the local error below a given threshold and
minimize the global computational effort.  This routine is meant to be
called from a user main program to perform the desired numerical
integration. This version of \taylorname{} (2.*) extends the
functionalities from \cite{JZ} by adding support of jet transport
\cite{GJJMZ}. I.e., generating code to integrate the variational
equations (of any order).

\smallskip 

Other features of \taylorname{} are:
\begin{enumerate}
\renewcommand{\theenumi}{\roman{enumi}}
    \item Automatic generator source code.
    \item High level of reusability.
    \item Flexibility in arithmetic definition, e.g.,
      multi-precision arithmetic.
    \item Thread safe code ({\sc new}).
%    \item Jet transport implementation ({\sc new}).
\end{enumerate}
\taylorname{} is a competitive software product in solving ODEs and it
is useful in computing:
\begin{itemize}
    \item high accuracy numerical ODE-solutions,
    \item directional variational equations,
    \item first order variational equations, and
    \item high-order variational equations.
\end{itemize}
\taylorname{} has direct applications in broad range of areas.  It
allows to compute, for instance, stability of equilibria, maximal
Lyapunov exponents, and high-order derivatives of Poincar\'e maps.

\subsubsection*{\taylorname{} highlights}

\begin{itemize}
    \item \taylorname{} is a high-order ODE-solver.
    \item \taylorname{} needs input of ODEs in its natural form.
    \item \taylorname{} admits user-defined arithmetic via a set of predefined
      \myfloat{} macros.
    \item \taylorname{} works with multi-precision arithmetic libraries.
    \item \taylorname{} uses an optimized step size control.
    \item \taylorname{} allows user-defined step size control.
    % \item \taylorname{} supports double complex variables but not complex time integration.
    \item \taylorname{} is OpenMP compatible. Declaration is required
      for thread dependent external parameters.
    \item \taylorname{} works with jet transport via its \myjet{}
      macros.
    % \item \taylorname{} assume \myjet{}'s coefficients as \myfloat{}.
\end{itemize}

\subsubsection*{What are the major differences with previous versions}
\taylorname{} (2.*) is compatible with version 1.* with the following
caveats:
\begin{enumerate}
\renewcommand{\labelenumi}{\theenumi.)}
    \item There is a new extra argument used for jet transport in the
      single step function. It needs to be set to \verb+NULL+ when jet
      transport is not needed.
    \item The header file now is ODE model dependent. Generation of
      header file requires an ODE model.
\end{enumerate}

\section{How to obtain \taylorname{}} \label{sec:download}
\taylorname{} is released under the GNU Public License (GPL).  It is
available on Github
\begin{tabbing}
    {\tt https://github.com/joang/taylor2-dist}
\end{tabbing}
\taylorname{} is also available on the web using the URLs
\begin{tabbing}
    {\tt http://www.math.utexas.edu/users/mzou/taylor/}\hspace*{5mm} \= (US)\\
    {\verb-http://www.maia.ub.es/~angel/taylor/-} \> (Europe)
\end{tabbing}



\section{How to install \taylorname{}} \label{sec:install}
\taylorname{} runs on Linux systems and it should compile and run on
other variant of Unixes. It has been tested on Linux and OsX.

\subsection{Installation from a repository}
\subsubsection*{Install instructions on Debian based systems}
\begin{itemize}
    \item {Install the Taylor repository key (running as the root
      user)
    
    
    \tt {wget -qO - https://web.ma.utexas.edu/repos/deb/taylor.gpg.key
      | apt-key add - } }
    \item{Add the Taylor repository 

    
    \tt{ echo "deb http://web.ma.utexas.edu/repos/deb focal main " >
      /etc/apt/sources.list.d/taylor.list }}
    \item{Install taylor 

    
    \tt{ apt-get update} 

    
    \tt{ apt-get install taylor}}
\end{itemize}

\subsubsection*{Install instructions on RedHat/Centos systems}
\begin{itemize}
    \item {Setup the repository (running as the root user)
    
    \tt{cd /etc/yum.repos.d}

    \tt{wget https://web.ma.utexas.edu/repos/rpm/taylor-redhat8.repo}
    }
    \item{Install taylor
    
    \tt{yum install taylor}
    
    }
\end{itemize}

\subsection{Installation from a Tarball}
After downloading the distribution {\tt taylor-x.y.z.tgz}, where {\tt
  x.y.z} is the version number, unpack the archive using the command
\begin{verbatim}
    tar xvzf taylor-x.y.z.tgz
\end{verbatim}
\noindent
or, if your version of {\tt tar} does not handle compressed files, you
can also use

\begin{verbatim}
    gzip -dc taylor-x.y.z.tgz | tar xvf - 
\end{verbatim}

\noindent
This will create a directory {\tt Taylor-x.y}.  Change to this
directory.

Now, to compile \taylorname{}, run {\tt make}.  It will produce the
executable {\tt taylor} in the current directory.  You need an ANSI C
compiler and {\tt lex/yacc} parser generator to compile
\taylorname{}. Using {\tt gcc} and {\tt flex/bison} is highly
recommended.

To install \taylorname{}, simply run
\begin{verbatim}
    make install 
\end{verbatim}



\section{What is jet transport in \taylorname{}} \label{sec:jettransport}
Jet transport refers to the time transport of what is called {\em jet}
through an ODE-solver. In \taylorname{}, a jet is implemented using
truncated formal power series. Given $m$ symbols $s_1, s_2, \ldots,
s_m$ and a fixed degree $d$, we represent the state variables as
\begin{equation}
    {\vec x} = {\vec x}_0+\sum_{0<k_1+k_2+\cdots+k_m \leq d} {\vec
      a}_{k_1k_2...k_m}{s_1}^{k_1}{s_2}^{k_2}\cdots{s_m}^{k_m}.
\end{equation}
$\vec x$ evolves using a Taylor ODE-solver method. During the process,
we replace the normal arithmetic on numbers with the same on truncated
power series. Different values of $m$ and $d$ give particular
interpretations, such as, when
\begin{itemize}
    \item $d = 0$: Usual Taylor method.
    \item $d=1$ and $m=1$: Directional first order variational equation.
    \item $m=1$: First order variational equation.
\end{itemize}

For efficiency, we implement several versions of arithmetic on
truncated power series although in the current version~\version{} the
default one is \verb+jet_tree+. These implementations are user
transparent unless you want to modify them. Here is the list.
\begin{itemize}
    \item \verb+jet1_1+ one symbol, degree one. The arithmetic is
      implemented using C macros.
    \item \verb+jet1+ one symbol, arbitrary degree. 
    \item \verb+jet_1+ degree one, arbitrary number of symbols.
    \item \verb+jet2+ two symbols, arbitrary degree.
    \item \verb+jet_2+ degree two, arbitrary number of symbols.
    \item \verb+jet_m+ a naive implementation of the general case,
      arbitrary number of symbols, and arbitrary degree. It works well
      when the jet size is below 1000 terms, i.e., whenever ${{d+m}
        \choose {d}} < 1000$.
    \item \verb+jet_tree+ a general library\footnote{based on {\tt
      http://www.maia.ub.es/dsg/param/chapter2.html}}.
\end{itemize}


\section{What is the syntax in \taylorname{}} \label{sec:syntax}

\subsection*{ODE declaration}
To use \taylorname{}, the first order of business is to prepare an
input file with the system of ODEs.  ODEs are specified by statements
like
\begin{verbatim}
    id = expr;
    diff(var, tvar) = expr;
\end{verbatim}
where {\tt tvar} is the time variable and {\tt expr} is a valid
mathematical expression made from numbers, the time variable, the
state variables, elementary functions $\sin$, $\cos$, $\tan$,
$\arctan$, $\sinh$, $\cosh$, $\tanh$, $\exp$, and $\log$, using the
four arithmetic operators and function composition. For example,
\begin{verbatim}
    a = log(1 + exp(-0.5));
    b = a + cos(0.1);
    c = a+b;
    ff = sin(x+t) * exp(-x*x);
    diff(x,t) = c * ff - tan(t);
\end{verbatim}
defines a single ODE.

\medskip 

\taylorname{} also understands {\tt if-else} expressions and
non-nested sums with static range. For example,
\begin{verbatim}
    ss = sum( i*sin(i * x)+ i *cos(i*t), i=1,10);
    diff(x,t) = ss;
    diff(y,t) = if(y>t) {  if(y>0.0) { y } else { 1-y }  }
                else    { y+t};
\end{verbatim}

\subsection*{Parameter declaration}
\taylorname{} allows to declare undefined parameters in the input file
following the syntax
\begin{verbatim}
    extern MY_FLOAT param;
\end{verbatim}
where {\tt extern} is a reserved \taylorname{}'s name, \myfloat{} is
the usual type (although it may be of other type, such as, {\tt
  double}, etc.), and {\tt param} refers to the parameter name.

\subsection*{Jet variables declaration}
Jet variables are declared using the statement
\begin{verbatim}
    jet [var_list|all] symbols [Number_of_symbols] degree [Degree];
\end{verbatim}
For example,
\begin{verbatim}
    jet x,y symbols 2 degree 2;
    jet all symbols 3 degree 1;
\end{verbatim}
In the first example, the variables {\tt x},{\tt y} will be treated as
polynomials of 2 symbols of degree up to 2.

In the second example, all the state variables will be polynomials of
3 symbols and degree 1. Note that if our system of ODEs consists of 3
equations, this is equivalent of integrating the ODEs along with its
first order variational equations.

\bigskip

If you only declare a subset of state variables as jet variables, you
need to make sure that
\begin{enumerate}
\renewcommand{\theenumi}{A\arabic{enumi}}
\renewcommand{\labelenumi}{\theenumi.)}
    \item\label{A1} the order of the variables must match that in the
      ODE specification, and they must appear before non-jet
      variables;
    \item\label{A2} the declared jet variables must be self contained,
      that is, all variables depends on any of the jet variables are
      also jet variables.
\end{enumerate}

Assumptions \ref{A1}--\ref{A2} force the system to have a coherent
output. For instance, the assumptions cover cases like systems of the
form
\begin{eqnarray*}
    \dot x &=& f(x,y), \\
    \dot y &=& g(y).
\end{eqnarray*}
If $y$ is a jet variable, then $x$ must be a jet variable due to
(\ref{A2}). However, it is allowed to consider $x$ a jet variable and
$y$ not; as far as $f$ is written first (\ref{A1}).

\section{How to run \taylorname{}}

Once the input file with the ODE system is ready, there are three main
steps in the construction of the \taylorname{} integrator.

\begin{enumerate}
\renewcommand{\theenumi}{\arabic{enumi}}
\renewcommand{\labelenumi}{\theenumi.)}
    \item\label{step1} We ask \taylorname{} to produce the code to
      compute the derivatives of the state variables, and the
      automatic step size (and order) control.
    \item\label{step2} We ask \taylorname{} to produce a header file
      that contains the definition of the data type, macros for basic arithmetic and forward declarations of
       API functions.
    \item\label{step3} We write a main driver function that repeatedly
      call the taylor integrator.
\end{enumerate}
Step~\ref{step1} generates a source code that is
arithmetic-independent, in the sense that the real numbers are
declared as \myfloat{} (type that is defined in step~\ref{step2}). All
the arithmetic operations have been replaced by C-macros. 

\smallskip

Step~\ref{step2} generates a file that must be included by the C-file
generated in Step~\ref{step1} and possibly other generated or custom
coded C files.

\medskip

Alternatively, \taylorname{} can also generate a sample main driver
function.  The above steps can be combined in a single one, by asking
\taylorname{} to output everything in a single file, that is, jet,
step size control, header, and a sample main.

\medskip

See Section~\ref{sec:examples} for illustrative examples or
Section~\ref{sec:advanced} for advanced details in steps~\ref{step1}
and \ref{step2}.

\section{How to use \taylorname{} by examples}\label{sec:examples}

The source code of examples in this manual is included in the \taylorname{} distribution.  If \taylorname{} is installed from a \verb+deb+ or \verb+rpm+ package, they're in \verb+/usr/share/taylor/+. If \taylorname{} is installed from source, they're in \verb+src/examples+ subdirectory.

\subsection{A first \taylorname{} contact} \label{sec:ex0}
\renewcommand{\inputfile}{\inputfiles{sec:ex0}}
\renewcommand{\odecfile}{\odecfiles{sec:ex0}}
\renewcommand{\odehfile}{\odehfiles{sec:ex0}}
\renewcommand{\mainfile}{\mainfiles{sec:ex0}}

Let \inputfile{} be a four lines ASCII file that specifies the famous
Lorenz equation.
\begin{code}[title={File: \inputfile{}}]
   RR  = 28.0;
   diff(x,t) = 10.0* (y - x);
   diff(y,t) = RR * x - x*z - y;
   diff(z,t) = x* y - 8.0* z /3.0;
\end{code}
After saving the file \inputfile{}, let us ask \taylorname{} to
generate a solver for us.

\subsubsection*{Single file output}
The simplest method is to ask \taylorname{} to generate everything in
a single file.  The command
\begin{command}
   taylor -o \odecfile{} \inputfile{}
\end{command}
produces a single file \odecfile{} ready to be compiled; 
\begin{command}
   gcc \odecfile{} -lm
\end{command}
If we run the binary ({\tt a.out}), the output looks like
\begin{verbatim}
    Enter Initial xx[0]: 0.03
    Enter Initial xx[1]: -0.02
    Enter Initial xx[2]: 0.15
    Enter start time: 0.0
    Enter stop time: 0.3
    Enter absolute error tolerance: 0.1e-16
    Enter relative error tolerance: 0.1e-16

0.03 -0.02 0.15 0
0.01836865  0.0061079  0.13463417  0.04051155
0.01802361  0.0269738  0.11958130  0.08501192
0.02707874  0.0550350  0.10484347  0.13448490
0.05034112  0.1086219  0.09053321  0.19018241
0.10579159  0.2304013  0.07720120  0.25359497
0.18285735  0.3986502  0.07016991  0.3
\end{verbatim}
The output of \verb+a.out+ are the values of the state variables, in
the order as they appear in the input file, plus the value of the time
variable. For our last example, each row of the output are values of
{\tt x}, {\tt y}, {\tt z}, and {\tt t}.

\subsubsection*{Standard use}
A more common use of \taylorname{} is when one needs to write its own
main driving procedure to call taylor stepper repeatedly, and perform
other operations on the output data. Hence, we normally need to ask
\taylorname{} to generate the header, the stepper, and the step
control functions in separate files.
\begin{command}
   taylor -name lrnz -o \odecfile{} -jet -step \inputfile{}
   taylor -name lrnz -o \odehfile{} -header \inputfile{}
\end{command}
The first line creates the file \odecfile{} ({\tt -o} flag) with the
code that computes the time-jet of derivatives ({\tt -jet} flag) and
the step size control ({\tt -step} flag); the ODE description is read
from the input file \inputfile{}. The flag {\tt -name}  tells
\taylorname{} the name we want to use for the stepper function; in this case the name is {\tt
  taylor\symbol{95}step\symbol{95}lrnz} (the string after the {\tt
  -name} flag is appended to the string {\tt
  taylor\symbol{95}step\symbol{95}} to get the name of this
function). The detailed description of the parameters of this function
is in Section~\ref{sec:or}.

The second line produces a header file (named \odehfile{}) needed to
compile \odecfile{}, that also contains the prototypes of the
functions in \odecfile{} (this is the reason for using again the flag
{\tt -name}) so the user may also want to include it to have these
calls properly declared. As we have not specified the kind of
arithmetic we want, this header file will use the standard double
precision of the computer.

\smallskip

Once we have the header and the stepper. We can write our main driving
routine to call the stepper to compute the orbit --this is similar to
the standard use of most numerical integrators, like Runge-Kutta or
Adams-Bashford.

\smallskip

As an example, let us ask \taylorname{} to create a very simple main
program for the Lorenz system,
\begin{command}
   taylor -name lrnz -o \mainfile{} -main\symbol{95}only \inputfile{}
\end{command}
Now we can compile and link these files,
\begin{command}
   gcc  \mainfile{} \odecfile{} -lm
\end{command}
to produce an executable. 


\subsubsection*{How to automatically generate a main program}
The default {\tt -main\symbol{95}only} flag generates a main program
that asks us to input initial values at run time. We can
specify the initial conditions, error tolerance, and stop conditions
in the input file.

\begin{center}
{\em We stress that this information is only used to produce the
  main() driving function.}
\end{center}

The syntax for specifying initial values is:
\begin{verbatim}
    initial_values = expr, expr, ..., expr;
\end{verbatim}
In the Lorentz input file, it would be
\begin{verbatim}
    initial_values = 2.03, 0.4, -0.5;
\end{verbatim}

For time step, error tolerance, and stop conditions, \taylorname{}
uses a few reserved variables (names).  They are:
\begin{verbatim}
    start_time = expr;               /* start time */
    stop_time = expr;                /* stop time: stop condition  */
    absolute_error_tolerance = expr; /* absolute error tolerance */
    relative_error_tolerance = expr; /* relative error tolerance */
    number_of_steps = expr;          /* stop condition */
\end{verbatim}
Here the right hand expressions must reduce to real constants.  The
lines \verb+stop_time+ and \verb+number_of_steps+ provide two
mechanisms to stop the integration. The solver will stop
when either condition is met.

\bigskip

Please be advised that expressions here are evaluated to double
precision values, then passed to the header macro
\verb+MakeMyFloatC(var,string_form,double_value)+.
When a supported multi-precision is used, this macro
uses a native conversion procedure to convert the number in string form to
a native multi-precision object.  \taylorname{} will pass
numbers in it's original form (i.e, not converted to double) to this macro. 
Expressions, however, will be evaluated to double first, its string form is then 
generated use the \verb+printf+ utility.

\bigskip
For example, we can add the following lines to 
\verb+lorenz.eq1+.
\begin{verbatim}
    initial_values= 0.03, -0.02, 0.15;
    start_time= 0.0;
    stop_time = 0.3;
    absolute_error_tolerance = 0.1e-16;
    relative_error_tolerance = 0.1e-16;
\end{verbatim}

\subsection{Extended precision arithmetic} \label{sec:ex1}
\renewcommand{\inputfile}{\inputfiles{sec:ex1}}
\renewcommand{\odecfile}{\odecfiles{sec:ex1}}
\renewcommand{\odehfile}{\odehfiles{sec:ex1}}
\renewcommand{\mainfile}{\mainfiles{sec:ex1}}

\taylorname{} has support for some extended precision arithmetic. For
instance, assume we want to generate a Taylor integrator for the
Lorenz example, using the GNU MPFR library.
\begin{command}
   taylor -mpfr -name lrnz -o \odecfile{} -jet -step \inputfile{}
   taylor -mpfr -name lrnz -o \odehfile{}  -header \inputfile{}
\end{command}
The flag {\tt -mpfr} instructs \taylorname{} to produce a header file
with {\bf mpfr} commands. As an example, we can ask \taylorname{} to
generate a (very simple) main program for this case,
\begin{command}
   taylor -mpfr -name lrnz -o \mainfile{} -main_only \inputfile{}
\end{command}
We stress that the {\bf mpfr} library is not included in the
\taylorname{} package. In what follows, we assume that it is already
installed in the computer and that {\bf mpfr} library is somewhere in
the default path used by your compiler to look for libraries,
otherwise you will need to tell the compiler ({\tt -L} flag for {\tt
  gcc}) where to find that library. Thus, to compile and line the
files, we use
\begin{command}
   gcc  \mainfile{} \odecfile{} -lmpfr -lm
\end{command}
to produce an executable.

\bigskip

\indent {\bf Important note:} Extended precision libraries usually
require some specific initializations that must be done by the main
program. The subroutines produced by \taylorname{} will crash or
produce wrong results if these initializations are not done
properly. We strongly suggest you to read the documentation that comes
with these libraries before using them.


\subsection{Using External Parameters} \label{sec:ex_params}
\renewcommand{\inputfile}{\inputfiles{sec:ex_params}}
\renewcommand{\odecfile}{\odecfiles{sec:ex_params}}
\renewcommand{\odehfile}{\odehfiles{sec:ex_params}}
\renewcommand{\mainfile}{\mainfiles{sec:ex_params}}

This example demonstrate the use of {\tt extern} variables, i.e.,
variable defined in somewhere other file that it must be find during
the linkage process.

In some cases, a vector field can depend on one or several parameters
and the user is interested in changing them at runtime. Moreover, for
vector fields that depends on lots of constants, e.g. power or Fourier
expansions, it is desirable to have a separate procedure to read in
those constants, rather than entering them by hand into the ODE
definitions. \taylorname{} understands external variables and external
arrays.  It treats them as constants when computing the taylor
coefficients.

\bigskip
{\tt extern} variables are declared by the {\tt extern} keyword.
Listed below is a short example.
\begin{code}[title={File: \inputfile{}}]
    extern MY_FLOAT e1, e2, coef[10], freq[10]; /* declare some external vars */
    
    diff(x,t) = e1 * y;
    diff(y,t) = -x + e2*sum( coef[i] * sin( freq[i] * t), i = 0, 9);
\end{code}
Let us save the above in \inputfile{}, and ask \taylorname{} to
generate a solver for us.
\begin{command}
    taylor -step -jet -o \odecfile{} -name perturbation \inputfile{}
    taylor -name perturbation -header -o \odehfile{} \inputfile{}
\end{command}
We will have to write a driver for our integrator. 

\begin{code}[title={File: \mainfile{}}]
    #include "^\odehfile{}^"

    /* these are the variables the vector fields depends on */
    MY_FLOAT e1, e2, coef[10], freq[10];
    int main(void)
    {
      MY_FLOAT xx[2], t;
      double h, abs_err, rel_err, h_return;
      double log10abs_err, log10rel_err, endtime;
      int i, nsteps = 1000, order=10, direction=1;
      int step_ctrl_method=2;
      /* read in e1, e2, coef[] and freq[]
       * here, we just assign them to some
       * values
       */
      e1 = e2 = 1.0;
      for(i = 0; i < 10; i++) {
        coef[i] = 1.0;
        freq[i] = 0.1*(double) i;
      }
      /* set initial condition */
      xx[0] = 0.1;
      xx[1] = 0.2;
      t = 0.0;
      /* control parameters */
      h= 0.001;
      abs_err = 1.0e-16;
      rel_err = 1.0e-16;
      log10abs_err = log10(abs_err);
      log10rel_err = log10(rel_err);
      endtime = 10.0;
      /* integrate 100 steps */
      h_return = h; /* thanks to Jason James */
      while( -- nsteps > 0 && h_return != 0.0 ) {
        /* do something with xx and t. We just print it */
        printf("%f %f %f\n", xx[0],xx[1],t);
        taylor_step_perturbation(&t, &xx[0], direction,
    			     step_ctrl_method,log10abs_err, log10rel_err,
    			     &endtime, &h_return, &order,NULL);
      }
      return 0;
    }
\end{code}

Note that the taylor steps require an initial time, the state
variable, the direction of integration (forward \verb|+1| or backward
\verb|-1|), a step control method, log10 absolute and relative errors,
and jet variables. Since the original system does not require jet
variables (see \inputfile{}), \mainfile{} sets to \verb+NULL+ the last
argument.

Now we can compile \odecfile{} and \mainfile{} and run the executable.
\begin{command}
   gcc \mainfile{} \odecfile{} -lm
   ./a.out
\end{command}

\subsection{First order directional derivative} \label{sec:ex_lyap}
\renewcommand{\inputfile}{\inputfiles{sec:ex_lyap}}
\renewcommand{\odecfile}{\odecfiles{sec:ex_lyap}}
\renewcommand{\odehfile}{\odehfiles{sec:ex_lyap}}
\renewcommand{\mainfile}{\mainfiles{sec:ex_lyap}}

Let us continue with the Lorenz model. In this example, we will
compute the positive maximal Lyaponov exponent along the famous Lorenz
attractor. Save the following text in \inputfile{}.
\begin{code}[title={File: \inputfile{}}]
    initialValues = 0.03,-0.02,0.15;
    absoluteErrorTolerance = 1.0E-16;   /* error tolerance for step control */
    relativeErrorTolerance = 1.0E-16;   /* error tolerance for step control */
    stopTime = 10000;                   /* stop time */
    startTime = 0.0;                    /* start time */
    
    /* ODE specification:  lorenz */
    RR = 28.0;
    x' = 10.0* (y - x);
    y' = RR* x - x*z - y;
    z' = x* y - 8.0* z /3.0;
    
    jet x,y,z symbols 1 deg 1;
    
    jestartTialValues x ="(0.03   1 )";
    jestartTialValues y ="(-0.02  0 )";
    jestartTialValues z ="(0.15   0 )";
\end{code}
Here we declare all our variables as jet variables of degree 1 in 1
symbol. We also specify the initial values for our jet variables. To
generate a header file \odehfile{} and a stepper file \odecfile{}.
\begin{command}
    taylor -header -o \odehfile{} \inputfile{}
    taylor -header_name \odehfile{} -jet -jhelper -step -o \odecfile{} \inputfile{}
\end{command}
Notice the \verb+-jhelper+ flag tells \taylorname{} to generate a few
jet IO helper functions.
%We will use one of them in our main driver, namely \verb+InputJetFromString+.

\bigskip
We need a main driving function. The following is modified from the
driver generated by \taylorname{}.
\begin{code}[title={File: \mainfile{}}]
    #include "^\odehfile{}^"
    
    int main(int argc, char *argv[])
    {
      int     i, j, order=20, itmp=0, direction = 1, nsteps = -1, counter=0;
      double  dstep, rtolerance, log10abs=-16, log10rel=-16;
      double  startT, stopT, nextT;
      double  xx[4], yy[4], zz[4], **jet;
      double  lyap=0.0, norm;
    
      MY_JET    *jetOut, jetIn[4];
    
      taylor_initialize_jet_library();
      for(i=0; i < 3; i++) InitJet(jetIn[i]);
      
      /* initialize jet vars --start */
      InputJetFromString(jetIn[0],"(0.03  1 )");
      InputJetFromString(jetIn[1],"(-0.02 0 )");
      InputJetFromString(jetIn[2],"(0.15  0 )");
      /* initialize jet vars --end */
    
      stopT = 10000;
      startT = 0;
      dstep=0.001;
    
      while(1)  {
        if(itmp != 0) {break;}
        if(startT >= stopT) { break;}
        itmp = taylor_step_lorenz2_eq(&startT, xx, direction, 1, 
                                      log10abs, log10rel, 
                                      &stopT, &nextT, &order, jetIn);
    
        if(++counter >= 1000) {  // we normalize the jet every 1000 steps                                                                             
          norm = 0;
          for(i=0; i < 3; i++)  norm += MY_JET_DATA(jetIn[i],1) * MY_JET_DATA(jetIn[i],1);  
    
          norm = sqrt(norm);
          lyap += log(norm);
    
          for(i=0; i < 3; i++) MY_JET_DATA(jetIn[i], 1) /= norm;
          counter = 0;
        }
      }  /* while */
      if(counter > 0) {
        for(i=0; i < 3; i++) norm += MY_JET_DATA(jetIn[i], 1) *  MY_JET_DATA(jetIn[i], 1); 
        norm = sqrt(norm);
        lyap += log(norm);
      }
    
      lyap /= 10000;
      fprintf(stdout, "The estimated Lyaponov exponent is: %f\n", lyap);
    
      exit(0);
    }
\end{code}
Now we can compile and run the code.
\begin{command}
    gcc \mainfile{} \odecfile{} -lm
    ./a.out
\end{command}
After a few seconds, it should report a result similar to the following.
\begin{verbatim}
    The estimated Lyapunov exponent is: 9.055203
\end{verbatim}%0.903542


\subsection{First order variational derivative} \label{sec:ex_varieq}
\renewcommand{\inputfile}{\inputfiles{sec:ex_varieq}}
\renewcommand{\odecfile}{\odecfiles{sec:ex_varieq}}
\renewcommand{\odehfile}{\odehfiles{sec:ex_varieq}}
\renewcommand{\mainfile}{\mainfiles{sec:ex_varieq}}

Following the Lorenz example, let us use \taylorname{} to solve the
first variational equations.  We also add a trivial equation to
include the variation with respect to a parameter. \inputfile{}
contains our Lorenz model, with an extra equation for the parameter
{\tt RR}.  The last line tells \taylorname{} to generate code to
compute the first variational flow.  It declares the jet variables (or
using the keyword {\tt all}), the number of symbols (4) and the degree
of jet (1).
\begin{code}[title={File: \inputfile{}}]
    x' = 10.0* (y - x);
    y' = RR* x - x*z - y;
    z' = x* y - 8.0* z /3.0;
    RR'= 0;
    
    jet x,y,z,RR symbols 4 deg 1;
\end{code}
To obtain the header and source of the ODE in \inputfile{} we can do
\begin{command}
    taylor -o \odehfile{} -header \inputfile{} -mpfr
    taylor -o \odecfile{} -jet -step -headername \odehfile{} \inputfile{} -jethelper -mpfr
\end{command}
Note that we used the flag {\tt -mpfr} to indicate that we want to use
the {\tt mpfr} multi-precision floating arithmetic. Note also that we
used the flag {\tt -name} to set a name of the ODE system, and used
{\tt -jethelper} flag to include jet IO helper functions.

\bigskip

The following \mainfile{} shows a main driving function. It is
arithmetic indepedent.  In other words, it works with code generated
by \taylorname{} with or without the {\tt -mpfr} flag.

\begin{code}[title={File: \mainfile{}}]
    #include "^\odehfile{}^"
        
    #define NS _NUMBER_OF_STATE_VARS_
    #define NJ _NUMBER_OF_JET_VARS_
    
    #ifdef _USE_MPFR_
    #define DIGITS_PRECISION 35
    #define STR(x) #x
    #define STR1(x) STR(x)
    #define JFMT "% ." STR1(DIGITS_PRECISION) "RNE"
    #else 
    #define DIGITS_PRECISION 16
    #define JFMT "% .14e"
    #endif
    
    int main(void)
    {
      int i, order, direction=+1, step_cntrl=2;
      double log10abs=-DIGITS_PRECISION, log10rel=-DIGITS_PRECISION;
      MY_FLOAT startT, nextT, stopT, x[NS];
      MY_JET xjet[NJ];
      
    #ifdef _USE_MPFR_
      mpfr_set_default_prec((int)(DIGITS_PRECISION*log2(10))+1);
    #endif
      
      taylor_initialize_jet_library();
      
      InitMyFloat(startT); InitMyFloat(nextT); InitMyFloat(stopT);
      for (i = 0; i < NS; i++) {InitMyFloat(x[i]);}    
      for (i = 0; i < NJ; i++) {taylor_initialize_jet_variable(&xjet[i]);}
      
      MakeMyFloatC(x[0],"0",0);
      MakeMyFloatC(x[1],"1",1);
      MakeMyFloatC(x[2],"0",0); 
      MakeMyFloatC(x[3],"28",28); /* parameter RR */
      
      taylor_make_identity_jets(xjet,x,NULL);
      
      MakeMyFloatC(startT,"0",0);
      MakeMyFloatC(stopT,"1",1);
      while (taylor_step_lorenz3_eq(&startT, x, direction, 
                                    step_cntrl, log10abs, log10rel, 
                                    &stopT, &nextT, &order, xjet) != 1) {}
      
      for (i = 0; i < NJ; i++) {
        printf("x%d=\n",i);
        taylor_output_jet(stdout, JFMT "\n", xjet[i]);
      }
      
      /* cleaning memory */
      for (i = 0; i < NJ; i++) {taylor_clear_jet_variable(&xjet[i]);}
      for (i = 0; i < NS; i++)  {ClearMyFloat(x[i]);}
      ClearMyFloat(stopT); ClearMyFloat(nextT); ClearMyFloat(startT); 
      taylor_clear_up_jet_library();
      return 0;
    } 
\end{code}
Finally, we can compile and run the code.
\begin{command}
    gcc \mainfile{} \odecfile{} -lgmp -lmpfr -lm
    ./a.out
\end{command}

% \subsection{A general \taylorname{} use} \label{sec:ex5}
% \renewcommand{\inputfile}{\inputfiles{sec:ex5}}
% \renewcommand{\odecfile}{\odecfiles{sec:ex5}}
% \renewcommand{\odehfile}{\odehfiles{sec:ex5}}
% \renewcommand{\mainfile}{\mainfiles{sec:ex5}}
% \jg{Just a code template of, e.g., Lorentz with all symbols deg 5 and using jet\_helper}


\section{How to become a \taylorname{} advanced user by examples}

\subsection{OpenMP compatibility} \label{sec:ex_omp}
\renewcommand{\inputfile}{\inputfiles{sec:ex_omp}}
\renewcommand{\odecfile}{\odecfiles{sec:ex_omp}}
\renewcommand{\odehfile}{\odehfiles{sec:ex_omp}}
\renewcommand{\mainfile}{\mainfiles{sec:ex_omp}}

\taylorname{} is openMP compatible provided that the jet variables
initialization are done properly in the driving routine.

\bigskip
If thread-dependent parameters are used in the ODE system, the user is
responsible to:
\begin{enumerate}
\renewcommand{\labelenumi}{\theenumi.)}
    \item add the \verb+#pragma omp threadprivate(param_name)+ in the
      driving routine, and
    \item append \verb+#pragma omp threadprivate(param_name)+ after
      the extern variable in the generated source file.
\end{enumerate}

We illustrate OpenMP use with an example.  Save the following ODE spec
in \inputfile{}.
\begin{code}[title={File: \inputfile{}}]
    extern MY_FLOAT RR,SS;
    x' = SS* (y - x);
    y' = RR* x - x*z - y;
    z' = x* y - 8.0* z /3.0;
    
    jet x,y,z symbols 3 deg 5;
\end{code}
We then generate the source and header files using \taylorname{},
\begin{command}
    taylor -o \odehfile{} -header \inputfile{}
    taylor -o \odecfile{} -jet -step -headername \odehfile{} \inputfile{} -jethelper
\end{command}
We assume parameter \verb+RR+ is thread-dependent, so we need to
modify \odecfile{}. Append the line
\begin{verbatim}
    extern MY_FLOAT RR;
\end{verbatim}
with 
\begin{verbatim}
    #pragma omp threadprivate(RR)
\end{verbatim}

\medskip
A driving file is provided in \mainfile{}. Notice that under the
assumption that \verb+RR+ is thread-dependent, the same \verb+pragma+
instruction is also added.

\begin{code}[title={File: \mainfile{}}]
    #include <omp.h>
    #include "^\odehfile{}^"
    #define NJ _NUMBER_OF_JET_VARS_
    #define NS _NUMBER_OF_STATE_VARS_
    MY_FLOAT RR, SS;
    #pragma omp threadprivate(RR)
    
    int main(int argc, char *argv[])
    {
      int k, j, ord;
      const int np = omp_get_max_threads();
      MY_FLOAT x[NS*np], te;
      MY_JET xjets[NJ*np];
    
      printf("np=%d threads\n", np);
    #pragma omp parallel
      taylor_initialize_jet_library();
      
      /* memory allocation */
      InitMyFloat(te); InitMyFloat(RR); InitMyFloat(SS);
      for (k = 0; k < NS*np; k++) {InitMyFloat(x[k]);}
      for (k = 0; k < NJ*np; k++) {InitJet(xjets[k]);}
      
      /* some initializations */
      MakeMyFloatC(SS,"10",10);
      MakeMyFloatC(te,"1",1);
      
      /* jets with identity matrix at first order */
      for (j = 0; j < np; j++) {taylor_make_identity_jets(xjets+NJ*j,x+NS*j,NULL);}
    
    #pragma omp parallel private(ord,k)
      {
        MY_FLOAT t,ht;
        InitMyFloat(t); InitMyFloat(ht);
    
        int tid = omp_get_thread_num();
        printf("tid=%d\n", tid);
        
        MakeMyFloatA(RR,tid+1);
        for (k = 0; k < NS; k++) {MakeMyFloatA(x[NS*tid+k],1/(tid+1));}
    
        MakeMyFloatC(t,"0",0);
        while (taylor_step_lorenz5_eq(&t, x+NS*tid, +1, 2, -16, -16,
                                      &te, &ht, &ord, xjets+NJ*tid) != 1) {}
    
        ClearMyFloat(ht); ClearMyFloat(t);
      } /* end parallel region */
      
      /* free memory */
      for (k = 0; k < NJ*np; k++) {ClearJet(xjets[k]);}
      for (k = 0; k < NS*np; k++) {ClearMyFloat(x[k]);}
      ClearMyFloat(SS); ClearMyFloat(RR); ClearMyFloat(te);
    #pragma omp parallel
      ClearUpJet();
      return 0;
    }
\end{code}
Finally, we can compile  and run the code
\begin{command}
    gcc \mainfile{} \odecfile{} -lm -fopenmp
    ./a.out
\end{command}

\subsection{Stability of equilibria}  \label{sec:ex_equilibria}
\renewcommand{\inputfile}{\inputfiles{sec:ex_equilibria}}
\renewcommand{\odecfile}{\odecfiles{sec:ex_equilibria}}
\renewcommand{\odehfile}{\odehfiles{sec:ex_equilibria}}
\renewcommand{\mainfile}{\mainfiles{sec:ex_equilibria}}

\taylorname{} can be used to study stability of equilibrium points by
computing its Jacobian and even bifurcations by computing high-order
derivatives.

Let us create the input ODE file \inputfile{} that codifies the famous
Lorenz model.
\begin{code}[title={File: \inputfile{}}]
    extern MY_FLOAT RR;
    x' = 10.0* (y - x);
    y' = RR* x - x*z - y;
    z' = x* y - 8.0* z /3.0;
    
    jet x,y,z symbols 3 deg 1;
\end{code}
and let us generate the source and header files using \taylorname{},
\begin{command}
    taylor -o \odehfile{} -header \inputfile{}
    taylor -o \odecfile{} -jet -step -headername \odehfile{} \inputfile{} -jethelper
\end{command}
The driving routine in \mainfile{} consists in assigning to the state
variable the equilibrium point and to the jet variables the
equilibrium points plus the identity matrix. After that, the
evaluation of the vector field gives us the differential at the
equilibrium point and from there one can compute the eigenvalues using
some linear algebra software.

Note that by increasing the degree, one can obtain higher derivatives
at the equilibrium point that are commonly required for bifurcation
studies.
\begin{code}[title={File: \mainfile{}}]
    #include "^\odehfile{}^"

    #define NJ _NUMBER_OF_JET_VARS_
    #define NS _NUMBER_OF_STATE_VARS_
    MY_FLOAT RR;
    
    int main(int argc, char *argv[])
    {
      int k, j;
      MY_FLOAT x[NS], t, A[NS*NS], *dtmp;
      MY_JET jetIn[NJ], **jetOut;
      
      taylor_initialize_jet_library();
      
      /* memory allocation */
      InitMyFloat(t);
      InitMyFloat(RR);
      for (k = 0; k < NS; k++) {InitMyFloat(x[k]);}
      for (k = 0; k < NJ; k++) {InitJet(jetIn[k]);}
      for (k = 0; k < NS*NS; k++) {InitMyFloat(A[k]);}
      
      /* equilibrium point */
      MakeMyFloatC(t,"0",0);
      MakeMyFloatC(RR,"28",28);
      for (k = 0; k < NS; k++) {MakeMyFloatC(x[k],"0",0);}
      
      /* jets with identity matrix at first order */
      taylor_make_identity_jets(jetIn,x,NULL);
      
      /* vector field evaluation */
      taylor_coefficients_lorenz5_eq_A(t, x, 1, 0, jetIn, &jetOut);
      
      for (k = 0; k < NS; k++) 
      {
        /* coefficients without the state variable */
        dtmp = taylor_convert_jet_to_array(jetOut[k][1],0);
        
        /* save the value in the matrix A*/
        for (j = 0; j < NS; j++) A[k*NS + j] = dtmp[j];   
      }
      
      /* print the differential at the equilibrium point */
      for (k = 0; k < NS*NS; k++) 
      {
        OutputMyFloat3(stdout,"% .5e ", A[k]);
        if ((k+1) % NS==0) printf("\n");
      }
      
      /* free memory */
      for (k = 0; k < NS*NS; k++) {ClearMyFloat(A[k]);}
      for (k = 0; k < NJ; k++) {ClearJet(jetIn[k]);}
      for (k = 0; k < NS; k++) {ClearMyFloat(x[k]);}
      ClearMyFloat(RR); ClearMyFloat(t);
      ClearUpJet();
      return 0;
    }
\end{code}


\section{Extra details for guru users} \label{sec:advanced}
This section is intended to provide extra details that an experienced
user may need, such as, \taylorname{} command line flags, output
routines, data types, API functions, stepsize controls, and driving
routines.


\subsection{Command Line Options}\label{sec:clo}

\taylorname{} supports the following command line options.
\begin{verbatim}
Usage: ./taylor 
  [-name ODE_NAME]
  [-o outfile] 
  [-long_double | -float128 | -mpfr | -mpfr_precision PRECISION ]
  [-main | -header | -jet | -jhelper | -jet_helper | -main_only] 
  [-step STEP_CONTROL_METHOD]
  [-jlib JET_LIBRARY]
  [-u | -userdefined] STEP_SIZE_FUNCTION_NAME ORDER_FUNCTION_NAME 
  [-sqrt]
  [-headername HEADER_FILE_NAME]
  [-debug] [-help] [-v]  file
\end{verbatim}

Let us explain them in detail.

\begin{itemize}
\item{ \verb+-name ODE_NAME+ 

This option specifies a name for the system of ODEs. The output
functions will have the specified name appended. For example, if we
run \taylorname{} with the option \verb+-name lorenz+, the output
procedures will be \verb+taylor_step_lorenz+ and
\verb+taylor_coefficients_lorenz+.  If name is not specified,
\taylorname{} appends the input filename (with non-alpha-numeric
characters replaced by \verb+_+) to its output procedure names. In the
case when input is the standard input, the word \verb+_NoName+ will be
used.}
\item{ \verb+-o outfile+ 

This option specifies an output file. If not specified, \taylorname{}
writes its output to the standard output.}

\noindent

\item{\verb+-mpfr+ 

This option tells \taylorname{} to use the the GNU MPFR library.  }

\item{\verb+-mpfr_precision PRECISION+

This flag is almost equivalent to {\tt -mpfr}; the only difference is
when a main() program is generated. If {\tt -mpfr} is used the main
program asks, at runtime, for the lenght (in bits) of the mantissa of
the {\bf mpfr} floating point types. If {\tt -mpfr\symbol{95}precision
  PRECISION} is used, the main program will set the precision to {\tt
  PRECISION} without prompting the user.}

\item{\verb+-float128+

This option tells \taylorname{} to use the GNU C \verb+__float128+
floating point arithmetic.  }

\item{\verb+-main+

Informs \taylorname{} to generate a very simple \verb+main()+ driving
routine. This option is equivalent to the options {\tt
  -main\symbol{95}only -jet -step 1}, so it produces a
``ready-to-run'' C file.  }

\item{\verb+-header+ 

This option tells \taylorname{} to output the header file. The header
file contains the definition of the \myfloat{} type (the type used to
declare real variables), macro definitions for arithmetic operations
and elementary mathematical function calls. In other words, this file
header file is responsible for the kind of arithmetic used for the
numerical integration.  Hence, the flag {\tt -header} must be combined
with one of the flags {\tt -mpfr}, {\tt -float128} to produce a header
file for the corresponding arithmetic. If none of these flags is
specified, the standard double precision arithmetic will be used.

Moreover, if the flag {\tt -name ODE\symbol{95}NAME} is also used, the
header file will also contain the prototypes for the main functions of
the Taylor integrator.}

\item{\verb+-jet+

This option asks \taylorname{} to generate only the code that computes
the taylor coefficients. The generated routine is

\begin{verbatim}
    MY_FLOAT **taylor_coefficients_ODE_NAME(
        MY_FLOAT t,  /* input: value of the time variable     */
        MY_FLOAT *x, /* input: value of the state variables   */
        int    order /* input: order of the taylor polynomial */
       )
\end{verbatim}

The code needs a header file (defining the macros for the arithmetic)
in order to be compiled into object code. The default header filename
is {\tt taylor.h}.  The header filename can be changed using
{\verb+-headername NAME+} (see below).  You can also use the
{\verb+-header+} option to include the necessary macros in the output
file.}

\item{\verb+-jhelper+

This option asks \taylorname{} to include jet IO helper functions in
the output.  When not combined with other output options,
\taylorname{} will only output the helper functions. For example,

\verb+ taylor -o helper.c -headername lorenz.h -jhelper lorenz.eq+

will save the helper functions in \verb+helper.c+. Please be advised
that
\begin{center} 
{\em the IO Helper functions should be generated only once per
  application.}
\end{center} 
Otherwise, the compiler will complain about multiple defined IO
helpers.  If you separate \taylorname{} generated code in files, it is
a good idea to keep the IO Helpers in its own file.  }
\item{\verb+-jlib JETLIBRARY+ 

This option select a jet library. By default, \taylorname{} uses the
\verb+jet_tree+ library. This option allows you to overwrite that with
a special purpose library. Possible value for \verb+JETLIBRARY+ are:
\begin{itemize}
    \item \verb+jet1_1+ one symbol, degree one. The arithmetic is
      implemented using C macros.
    \item \verb+jet1+  one symbol, arbitrary degree. 
    \item \verb+jet_1+ degree one, arbitrary number of symbols.
    \item \verb+jet2+ two symbols, arbitrary degree.
    \item \verb+jet_2+ degree two, arbitrary number of symbols.
    \item \verb+jet_m+ an naive implementation of the general case,
      arbitrary number of symbols, arbitrary degree. This
      implementation works well when the jet size is below 1000 terms,
      i.e., when ${{d+m} \choose {d}} < 1000$.
    \item \verb+jet_tree+ a general library\footnote{based on {\tt
      http://www.maia.ub.es/dsg/param/chapter2.html}}. This is the
      default.
\end{itemize}
}
\item{\verb+-main_only+

This option asks \taylorname{} to generate only the \verb+main()+
driving routine. It is useful when you want to separate different
modules in different files. The main driving routine has to be linked
with the step size control procedure and the jet derivative procedure
to run.  }

\item{\verb+-step STEP_SIZE_CONTROL_METHOD+

This option asks \taylorname{} to generate only the order and step
size control code supplied by the package. If combined with the {\tt
  -main} or {\tt -main\symbol{95}only} flags, the value {\tt
  STEP\symbol{95}SIZE\symbol{95}CONTROL\symbol{95}METHOD} is used in
the main program to specify the step size control. The values of {\tt
  STEP\symbol{95}SIZE\symbol{95}CONTROL\symbol{95}METHOD} can be 0
(fixed step and degree), 1, 2, and -1 (user defined step size control;
in this case you have to code your own step size and degree
control). If the flags {\tt -main} and {\tt -main\symbol{95}only} are
not used, this value is ignored.

The generated procedure is also the main call to the numerical
integrator:
\begin{verbatim}
    int taylor_step_ODE_NAME(MY_FLOAT *time,
                             MY_FLOAT *xvars,
                             int      direction,
                             int      step_ctrl_method,
                             double   log10abserr,
                             double   log10relerr,
                             MY_FLOAT *endtime,
                             MY_FLOAT *stepused,
                             int      *order,
                             MY_JET   *jetInOut)
\end{verbatim}
This code needs the header file to be compiled (see the remarks
above). Given an initial condition ({\tt time},{\tt xvars},{\tt
  jetInOut}), this function computes a new point on the corresponding
orbit with the jet variables transported. The meaning of the
parameters is explained in Section~\ref{sec:or}.  }

\item{\verb+-userdefined STEP_SIZE_FUNCTION_NAME  ORDER_FUNCTION_NAME+

This flag specifies the names of your own step size and order control
functions. Then, the code produced with the flag {\tt -step} includes
the calls to your control functions; to use them, you must set {\tt
  step\symbol{95}ctrl\symbol{95}method} to 3 (see
Section~\ref{sec:tnu}).

For more details (like the parameters for these control functions)
look at the source code produced by the {\tt -step} flag.  }

\item{\verb+-sqrt+

This option tells \taylorname{} to use the function \verb+sqrt+
instead of \verb+pow+ when evaluating terms like $(x+y)^{-{3\over
    2}}$.  The use of {\tt sqrt} instead of {\tt pow} produces code
that runs faster.}

\item{\verb+-headername HEADER_FILE_NAME+

When \taylorname{} generates the code for the jet and/or step size
control, it assumes that the header file will be naed {\tt taylor.h}.
This flag forces \taylorname{} to change the name of the file to be
included by the jet and/or step size control procedures to the new
name {\tt HEADER\symbol{95}FILE\symbol{95}NAME}. Of course, the user
is then responsible for creating such a header file by combining the
flags {\tt -o HEADER\symbol{95}FILE\symbol{95}NAME} and {\tt
  -header}. For instance,
\begin{verbatim}
    taylor -name lz -o l.c -jet -step -headername l.h lorenz.eq1
\end{verbatim}
stores the code for the jet of derivatives and step size control in
the file {\tt l.c}. Moreover, {\tt l.c} includes the header file {\tt
  l.h}. This file has to be created separately:
\begin{verbatim}
    taylor -name lz -o l.h -header lorenz.eq1
\end{verbatim}
}

\item{\verb+-debug+ or \verb+-v+ 

Print some debug info to \verb+stderr+.}

\item{\verb+-help+ (or \verb+-h+)

Print a short help message.}

\end{itemize}

\noindent
The default options are set to produce a full C program, using the
standard double precision of the computer:
\begin{verbatim}
   -main_only -header -jet -step 1
\end{verbatim}

\subsection{The Output Routines}\label{sec:or}
\taylorname{} outputs two main procedures. The first one is the main
call for the integrator and the second one is a function that computes
the jet of derivatives. For details on some other routines generated
by \taylorname{} (like degree or step size control), see the comments
in the generated source code.

\subsubsection*{The numerical integrator}\label{sec:tnu}
Its prototype is:
\begin{verbatim}
   int taylor_step_ODE_NAME(MY_FLOAT *time,
                            MY_FLOAT *xvars,
                            int      direction,
                            int      step_ctrl_method,
                            double   log10abserr,
                            double   log10relerr,
                            MY_FLOAT *endtime,
                            MY_FLOAT *stepused,
                            int      *order,
                            MY_JET   *jetOut);
\end{verbatim}
The function \verb+taylor_step_ODE_NAME+ does one step of numerical
integration of the given system of ODEs, using the control parameters
passed to it. It returns 1 if \verb+endtime+ is reached, 0 otherwise.

\bigskip

\noindent{\bf Parameters:}
\begin{itemize}
\item {\tt time}\newline
  on input:  time of the initial condition\newline
  on output: new time
\item {\tt xvars}\newline
  on input:  initial condition\newline
  on output: new condition, corresponding to the (output) time
\item {\tt direction}\newline
  flag to integrate forward or backwards.\newline
  \hspace*{5mm} \phantom{-}1: forward\newline
  \hspace*{5mm} -1: backwards\newline
  Note: this flag is ignored if {\tt
    step\symbol{95}ctrl\symbol{95}method} is set to 0.
\item {\tt step\symbol{95}ctrl\symbol{95}method}\newline flag for the
  step size control. Its possible values are:
  \begin{itemize}
  \item[\phantom{-}0:] no step size control, so the step and order are
    provided by the user. The parameter \verb+stepused+ is used as
    step, and the parameter \verb+order+ (see below) is used as the
    order.
  \item[\phantom{-}1:] standard stepsize control. It uses an
    approximation to the optimal order and to the radius of
    convergence of the series to approximate the ``optimal'' step
    size. It tries to keep the absolute and relative errors below the
    given values. See the paper \cite{JZ} for more details.
  \item[\phantom{-}2:] as 1, but adding an extra condition on the
    stepsize $h$: the terms of the series -- after being multiplied by
    the suitable power of $h$ -- cannot grow.
  \item[-1:] user defined stepsize control. The code has to be
    included in the source routine called {\tt
      compute\symbol{95}timestep\symbol{95}user\symbol{95}defined}
    (see the code). The user must also include code for the selection
    of degree, in the function {\tt
      compute\symbol{95}order\symbol{95}user\symbol{95}defined}.
  \end{itemize}
\item {\tt log10abserr}\newline
  decimal log of the absolute accuracy required.
\item {\tt log10relerr}\newline
  decimal log of the relative accuracy required.
\item {\tt endtime}\newline
  if {\tt NULL}, it is ignored. \\if {\tt step\symbol{95}ctrl\symbol{95}method} is set to
  0, it is also ignored. \\otherwise, if next step is going to be
  outside {\tt endtime}, reduce the step size so that the new time
  {\tt time} is exactly {\tt endtime} (in that case, the function
  returns 1).
\item {\tt ht}\newline
  on input: ignored/used as a time step (see parameter {\tt
  step\symbol{95}ctl\symbol{95}method})\newline
  on output: time step used
\item {\tt order}
 \begin{description}
  \item input: this parameter is only used if {\tt
    step\symbol{95}ctrl\symbol{95}method} is 0, or if you add the
    proper code for the case {\tt
      step\symbol{95}ctrl\symbol{95}method}=3.\newline If {\tt
      step\symbol{95}ctrl\symbol{95}method} is 0, its possible values
    are:\newline $<2$: the program will select degree 2,\newline $\ge
    2$: the program will use this degree.
  \item output: degree used.
  \end{description}
\item {\tt jetInOut}
\begin{description}
\item on input: the value of jet variables
\item on output: the new value of jet variable at the output time {\tt ti}.
\end{description}
\end{itemize}

\noindent {\bf Returned value:}
\begin{itemize}
\item 0: ok.
\item 1: ok, and {\tt time}={\tt endtime}.
\end{itemize}


\subsubsection*{The jet of derivatives}
Its prototype is
\begin{verbatim}
    MY_FLOAT **taylor_coefficients_ODE_NAME(MY_FLOAT t,
                                            MY_FLOAT *x,
                                            int order);
\end{verbatim}
\verb+taylor_coefficients_ODE_NAME+ returns a {\bf static} two
dimensional array. The rows are the Taylor coefficients of the state
variables.

\vspace{3mm}
\noindent{{\bf{Parameters}}}
\begin{itemize}
  \item{\verb+t+: value of the time variable. It is used only when the
    system of ODEs is nonautonomous.}
  \item{\verb+x+: value of the state variables.}
  \item{\verb+order+: degree of Taylor polynomial.}
\end{itemize}

If you want to compute several jets at the same point but with
increasing orders, then you should consider using the call
\begin{verbatim}
    MY_FLOAT **taylor_coefficients_ODE_NAMEA(MY_FLOAT t,
                                             MY_FLOAT *x,
                                             int order,
                                             int rflag,
                                             MY_JET *jetIn, 
                                             MY_JET ***jetOut)
\end{verbatim}
(note the ``{\tt A}'' at the end of the name). The first three
parameters have the same meaning as before, and the meaning of the
fourth one is:
\begin{description}
  \item 0: the jet is computed from order 1 to order {\tt order}.
  \item 1: the jet is computed starting from the final order of the
    last call, up to {\tt order}.
\end{description}

Care must be exercised if you invoke this routine with {\tt
  rflag=1}. Indeed, if you modify the Taylor coefficients and/or the
base point, you need to restore them before the next call.


\subsection{The data type \myfloat{}} \label{sec:myfloat}

\myfloat{} is a customizable floating point data type.  Arithmetic on
\myfloat{} is defined through a set of C macros. \taylorname{}
currently supports the following floating point data types.
\begin{itemize}
    \item \verb+double+.  The standard C \verb+double+.  This is the
      default.
    % \item \verb+double complex+. A complex double precision arithmetic based on the C standard \verb+complex.h+. 
    \item \verb+long double+. The C \verb+long double+ type.
    \item \verb+__float128+. The GNU C 128 bit floating point
      type\footnote{\texttt{https://gcc.gnu.org/onlinedocs/libquadmath/index.html}}.
    \item \verb+mpfr_t+. An arbitary precision floating-point data
      type from {\tt mpfr}
      library\footnote{\texttt{https://www.mpfr.org}}.
    \item \verb+gmp_t+. An arbitrary precision floating point data
      type from {\tt gmp}
      library\footnote{\texttt{https://gmplib.org}}.
\end{itemize}
These data types can be selected using \taylorname{} command line
flags, that is, \verb+-long_double+, \verb+-float128+, \verb+-gmp+ or
\verb+-gmp_precision NBITS+ and \verb+-mpfr+ or
\verb+-mpfr_precision NBITS+, see Section~\ref{sec:clo}.

\bigskip
To extend \taylorname{} to support a custom floating point data type,
one needs to redefine the set of C macros \taylorname{} calls for. See
{\tt{mpfrheader.h}} in the source code for reference.


\subsection{The data type \myjet{}} \label{sec:myjet}
\myjet{} is a new data type that encodes truncated multivariate
polynomial through an array of monomial coefficients. The exact
implementation is library dependent and hence is opaque to the end
user. In all implementations, monomial coefficients are stored as an
array of \myscal{}s. A C macro \verb+MY_JET_DATA(jet,i)+ is provided
to access the $i^{{\tt th}}$ monomial coefficient, arranged in
``graded lexicographical order''.  In the current version of
\taylorname{}, \myscal{} is just an alias of \myfloat{}. This may
change in future releases.

\subsubsection*{API functions related to \myjet{}} \label{sec:myjetapi}
For your convenience, \taylorname{} includes a few API functions
pertained to Jet Transport. Most of these functions are IO
helpers. These functions will be generated when the \verb+-jhelper+
command line option is passed to \taylorname{}.

\bigskip
The API functions use size related constants or global variables
generated by \taylorname{}. Those variables are computed from the
number of symbols and the degree from jet declaration line in ode
specification.  It is assumed that those variables stay unchanged in
an application. Thus, if you use \taylorname{} to integrate multiple
ODEs in the same program, the jet must be declared the same way in
all, i.e, use the same degree and number of symbols in all ODE
specifications.

\bigskip
Some of the API functions take both \verb+MY_FLOAT+ and \verb+double+
arguments to set values of a \myjet{} variable.  For high accuracy
computations, it is advised that you use the \verb+MY_FLOAT+ argument,
as the conversion of double to \verb+MY_FLOAT+ is not precise.

\begin{itemize}
\item {\verb+int taylor_make_jet(MY_JET a, MY_FLOAT *myfloats, double *values)+  \\  
    This function fills in the data array of a \myjet{} with either
    \verb+myfloats+ or \verb+values+. Caller is responsible to make
    sure there are enough values in \verb+myfloats+ or
    \verb+values+. It should include the coefficients of all monomials
    for the current model, including the constant term.}
    
\item{\verb+int taylor_make_identity_jets(MY_JET *inOut, MY_FLOAT *myfloats, double *values)+ \\
    This function fills the array \verb+inOut+ with an identity
    matrix. If either \verb+myfloats+ or \verb+values+ is not NULL, it
    fills the constant terms in \verb+inOut+ with elements in
    \verb+myfloats+ or \verb+values+.  This function uses the global
    variable \verb+_number_of_jet_vars_+, \verb+_number_of_symbols_+
    and \verb+_size_of_jet_variable_+.  Please note, this function
    only fills the first \verb+_number_of_symbols_+ or
    \verb+_number_of_jet_vars+ rows of \verb+inOut+, whichever is
    smaller. The constant terms are filled in full, if provided.}
    
\item{\verb+int taylor_make_unit_jet(MY_JET a, int idx, MY_FLOAT *myfloat, double *value)+\\
    This function assigns 1 to the coefficient of the \verb+idx+th
    symbol in \verb+a+, assign 0 to other coefficients. If either
    \verb+myfloat+ or \verb+value+ is not NULL, its value is assigned
    to the constant term in \verb+a+.  }
    
\item{\verb+int taylor_set_jet(MY_JET a, MY_FLOAT *myfloats, double *values, int include_state)+\\
    This function fills in the jet variable a with values supplied in
    \verb+myfloats+ or \verb+values+. \verb+include_state+ signals if
    the constant term is included in \verb+myfloats+ or
    \verb+values+. If not, 0 will be assigned to the constant term in
    a. }
    
\item{\verb+int taylor_input_jet_from_stdin(MY_JET a, int idx)+\\
    This function asks the user to fill in a jet variable from stdin.
    \verb+idx+ signals which state variable \verb+a+ is associated
    with.  This function prints out a prompt that includes all
    monomials.}
    
\item{\verb+int taylor_input_jet_from_string(MY_JET a, const char *str)+\\
    This function fills in the jet variable \verb+a+ with values from
    a string. It is an easy method to fill in a jet variable. For
    example, \verb+InputJetFromString(jetIn[0],"(0.03 1 0 0 0 )");+.}
    
\item \verb+int taylor_output_jet(FILE *file, char *fmt, MY_JET a)+\\
    This function output a jet variable to \verb+file+, using the
    supplied format to format each element. Please note, some
    \myfloat{} type may require special flags in the format
    string. For example, \verb+long double+ requires a \verb+L+
    specifier, like \verb+%.18Lf+.

\item{\verb+void taylor_initialize_jet_library()+\\    
    This function initializes the JET library. }
    
\item{\verb+void taylor_initialize_jet_variable(MY_JET *jet)+\\
    This function initializes the jet variable \verb+jet+. All jet
    variables need to be initialized before use.}
    
\item{\verb+void taylor_clear_jet_variable(MY_JET *jet)+\\
  This function clears the storage space allocated for a jet variable
  \verb+jet+.}
  
\item{\verb+int taylor_set_jet_variable_degree(int d)+\\
    This function sets the working degree of all jet variables.  The
    degree must be smaller than the maximum degree set in the ODE
    specification.  This function returns the current working degree.}
    
\item{\verb+void taylor_jet_reduce(MY_JET a, double *v)+\\
    This function reduces a jet variable \verb+a+ to a number, using
    the values in \verb+v+ to substitute the symbols in the
    corresponding position.}
    
\item{\verb+const char **taylor_get_variable_names()+\\
    This function returns the list of names for all state variables,
    in the order as they appear in the model specification.}
    
\item{\verb+const char **taylor_get_jet_monomials()+\\
    This function returns the list of monomials in 'graded
    lexicographic' order for the current model. Because generating the
    list of monomials is an expensive operation. This function will
    return an empty list when the total number of monomials exceeds
    1000.}
\end{itemize}


\subsubsection*{Accessing elements in \myjet{}} \label{sec:ex_myjetaccess}
\renewcommand{\inputfile}{\inputfiles{sec:ex_myjetaccess}}
\renewcommand{\odecfile}{\odecfiles{sec:ex_myjetaccess}}
\renewcommand{\odehfile}{\odehfiles{sec:ex_myjetaccess}}
\renewcommand{\mainfile}{\mainfiles{sec:ex_myjetaccess}}

Internally, \myjet{} stores monomial coefficients in an array of
\myfloat{}s. The order of the coefficients differs by
implementation. A macro \verb+MY_JET_DATA(jet,idx)+ is provided for
accessing these coefficients in ``graded lexicographic'' order.  That
is, the monomials are ordered by total degree first, monomials of the
same total degree are ordered lexicographically.  For example, for
degree 2 with 3 symbols, the "graded lexicographic" order of the
monomials are
\[
1, s_1, s_2, s_3, s_1^2, s_1s_2, s_1s_3, s_2^2, s_2s_3, s_3^2
\]
\verb+MY_JET_DATA(jet,6)+ refers the coefficient of $s_1s_3$.  This
macro can be used as lvaue, i.e., to assign/change values of the
monomial coefficients.

\vspace{4mm}

The following example demonstrates the use \verb+MY_JET_DATA+.

\vspace{4mm}
Save the following ODE specification in \inputfile{}
\begin{code}[title={File: \inputfile{}}]
    x' = -y + x *( 1-x*x -y*y);
    y' =  x + y *( 1-x*x -y*y);
    r' = r*(1-r*r);
    
    jet x,y  symbols 2 deg 2;
\end{code}
and save the following code to \mainfile{}.
\begin{code}[title={File: \mainfile{}}]
    #include "^\odehfile{}^"
    
    /**
     *  This example demostrates how to access MY_JET storage array directly.
     */
    
    #define NN  _NUMBER_OF_STATE_VARS_
    #define JJ  _NUMBER_OF_JET_VARS_ 
    
    double initial_values[] = {0.03,-0.02,0.15,0.1,-0.1,0.1,0.2,-0.2,0.2,0.1};
    double T0 = 0.0, T1 = 0.2;
     
    int main(int argc, char *argv[])
    {
    	 int       i, j, order=20, itmp=0, direction = 1, nsteps = -1;
    	 double    dstep, log10abs=-16, log10rel=-16;
    	 MY_FLOAT  startT, stopT, nextT, xx[NN];
    	 MY_JET    jetIn[JJ];
    	 
    #ifdef _USE_MPFR_	 
    	 mpfr_set_default_prec(256);
    #endif
    #ifdef _USE_GMP_	 
    	 mpf_set_default_prec(256);
    #endif	 
    	 taylor_initialize_jet_library();
    	 
    	 InitMyFloat(startT);
    	 InitMyFloat(stopT);
    	 InitMyFloat(nextT);
    	 
    	 for(i = 0; i<NN; i++) {InitMyFloat(xx[i]);}
    	 for(i = 0; i<JJ; i++) {InitJet(jetIn[i]);}
    
    
    	 dstep=0.001; /* only nedeed when step_ctrl_method==0 (see manual) */
    
    	 taylor_make_identity_jets(jetIn, NULL, NULL);
    
    	 MakeMyFloatA(stopT, T1);
    	 MakeMyFloatA(startT, T0);	 
    	 order=20;
    	 itmp = 0;
    
    	 for(i =0; i< NN; i++) {
    	   MakeMyFloatA(xx[i],  initial_values[i]);
    	   // here we assign value to the constant term in jetIn[i]
    	   AssignMyFloat(MY_JET_DATA(jetIn[i], 0),  xx[i]);	   	   
    	 }
    
    	 // print monomial names
    	 {
    	   char **monomials = taylor_get_jet_monomials();
    	   fprintf(stdout,"%-18s", "     1");
    	   i = 0;
    	   while(monomials[i] != NULL) {
    	     fprintf(stdout, "%-13s", monomials[i]);
    	     i++;
    	   }
    	   fprintf(stdout,"\n");
    	 }
    	 
    	 while(1)  {
    	   for(i = 0; i < JJ; i++) {
    	     char **var_names = taylor_get_variable_names();
    	     // here we print all monomial coefficients	   	     
    	     fprintf(stdout, "%s: ", var_names[i]);
    	     for(j = 0; j < _SIZE_OF_JET_VAR_; j++) {
    #ifdef _USE_MPFR_	 	       
    	       fprintf(stdout, "%12.8f ", mpfr_get_d(MY_JET_DATA(jetIn[i],j),GMP_RNDN));
    #else
    #ifdef _USE_GMP_
    	       fprintf(stdout, "%12.8f ", mpf_get_d(MY_JET_DATA(jetIn[i],j),GMP_RNDN));	       
    #else
    	       fprintf(stdout, "%12.8f ", MY_JET_DATA(jetIn[i],j));	       
    #endif
    #endif	       
    	     }
    	     fprintf(stdout, "\n");	     
    	   }
    	   fprintf(stdout, "\n");	     	   
    	   
    	   if(itmp != 0) {break;} 
    	   if(MyFloatA_GE_B(startT,stopT)) { break;}
    	   
    	   itmp = taylor_step_jdata( &startT, xx, direction, 1, 
    	                             log10abs, log10rel, 
    	                             &stopT, &nextT, &order, jetIn);
    
    	 }  /* while */
    	 
    	 exit(0); 
     }   
\end{code}
Now generate the required header \odehfile{} and a taylor stepper \odecfile{}.
\begin{command}
   taylor -header -name jdata -o \odehfile{} \inputfile{} 
   taylor -jet -jhelper -name jdata -step -header_name \odehfile{} \inputfile{} -o \odecfile{}
\end{command}
Finally compile the code and run
\begin{command}
    cc -g \odecfile{} \mainfile{} -lm
    ./a.out
\end{command}
You'll get a nice printout of elements in \myjet{} like the following.
\begin{verbatim}
     1            s1           s2           s1^2         s1s2         s2^2         
x:   0.03000000   1.00000000   0.00000000   0.00000000   0.00000000   0.00000000 
y:  -0.02000000   0.00000000   1.00000000   0.00000000   0.00000000   0.00000000 

x:   0.04075176   1.19607263  -0.24217713  -0.02764699   0.01532271  -0.01239364 
y:  -0.01665614   0.24282333   1.19650979   0.00051102  -0.01525349   0.01585447 
\end{verbatim}


\subsection{Write a Driving Routine}\label{sec:ex_driving}
\renewcommand{\inputfile}{\inputfiles{sec:ex_driving}}
\renewcommand{\odecfile}{\odecfiles{sec:ex_driving}}
\renewcommand{\odehfile}{\odehfiles{sec:ex_driving}}
% \renewcommand{\mainfile}{\mainfiles{sec:ex_driving}}
The main driving routine produced by the {\tt -main} flag of
\taylorname{} is rather simple, it just keeps on integrating the
system and print out the solution along the way.  This may be enough
for some tasks, but it is definitely too primitive for real
applications.  In this section, we provide two sample driving
routines.  These examples demonstrate what you need to do to write
your own driving routes. The input files are provided in the doc
subdirectory in the \taylorname{} distribution.

We first ask \taylorname{} to generate a integrator and a header file
for us.
\begin{command}
   taylor -o \odecfile{} -jet -step -name lorenz \inputfile{}
   taylor -o \odehfile{} -header \inputfile{}
\end{command}
The first command will produce a file \odecfile{} with no driving
routine in it.  This file will be compiled and linked with our main
driving routine.  The second command generates the header file
\odehfile{}. It is needed in \odecfile{} and our main driving
function.

\subsubsection*{Using the Supplied Integrator}

Our first example is very similar to the driving routine generated by
\taylorname{}.  It uses the one step integrator provided by
\taylorname{}.

\begin{code}[title={File {\tt main1.c}}]
    #include <stdio.h>
    #include <math.h>
    #include "^\odehfile{}^"
    int main(int argc, char *argv[])
    {
       MY_FLOAT  xx[3], t;
       double    h, h_return, log10abs_err, log10rel_err, endtime;
       int       nsteps = 100, step_ctrl_method = 2, direction = 1;
       int       order = 10;
       /* set initial conditions */
       xx[0] = 0.1;
       xx[1] = 0.2;
       xx[2] = 0.3;
       t     = 0.0;
       /* control parameters       */
       h= 0.001;
       log10abs_err = -16 /* i.e. 1.0e-16 absolute error */
       log10rel_err = -16 /* i.e. 1.0e-16 relative error */
       endtime = 10.0;
    
       /* integrate 100 steps */
       while( -- nsteps > 0 && h_return != 0) {
          /* do something with xx and t. We just print it */
          printf("%f %f %f %f\n", xx[0],xx[1],xx[2],t);
          taylor_step_lorenz(&t, &xx[0], direction,
                             step_ctrl_method,log10abs_err, log10rel_err, 
                             &endtime, &h_return, &order);
       }
       return 0;
    }
\end{code}

\noindent
After saving the code in \verb+main1.c+, you can compile them using
the command
\begin{command}
   gcc \odecfile{} main1.c -lm
\end{command}
and run the executable \verb+a.out+ as before.


\subsubsection*{Writing Your Own Driver}

This example provides a skeleton for writing your own one step
integrator.

\begin{code}[title={File {\tt main2.c}}]
    #include <stdio.h>
    #include <math.h>
    #include "^\odehfile{}^"
    
    MY_FLOAT **taylor_coefficients_lorenz(MY_FLOAT, MY_FLOAT *, int);
    
    int main(int argc, char *argv[])
    {
       MY_FLOAT  xx[3], tmp[3], t, **coef;
       int       j, order=20, nsteps = 100;
       double    step_size;
       /* set initiaial conditions */
       xx[0] = 0.1;
       xx[1] = 0.2;
       xx[2] = 0.3;
       t     = 0.0;
       /* control parameters   */
       step_size= 0.1;
    
       /* integrate 100 steps */
       while( -- nsteps > 0) {
          /* do something with xx and t. We just print it */
          printf("%f %f %f %f\n", xx[0], xx[1], xx[2], t);
    
          /* compute the taylor coefficients */
          coef = taylor_coefficients_lorenz(t, xx, order);
    
          /* now we have the taylor coefficients in coef,
           * we can analyze them and choose a best step size.
           * Here we just integrate use the given stepsize.
           */
           tmp[0] = tmp[1] = tmp[2] = 0.0;
           for(j=order; j>0; j--) /* sum up the taylor polynomial */
             { 
               tmp[0] = (tmp[0] + coef[0][j])* step_size;
               tmp[1] = (tmp[1] + coef[1][j])* step_size;
               tmp[2] = (tmp[2] + coef[2][j])* step_size;
             }
            /* advance one step */
            xx[0] = xx[0] + tmp[0];
            xx[1] = xx[1] + tmp[1];
            xx[2] = xx[2] + tmp[2];
            t += step_size; /* advance time */
         }
        return 0;
    }
\end{code}

\appendix

\section{The Taylor method}
Taylor method is one of the best known one step method for solving
ordinary differential equations numerically. The idea is to advance
the solution using a truncated Taylor expansion of the variables about
the current solution.  Let
\begin{eqnarray}
\label{eq:ivp}
{\bf y}' = f(t, {\bf y})\qquad {\bf y}(t_0) = {\bf y}_0
\end{eqnarray}
be an initial value problem and let $h$ be the integration step. To
find ${\bf y}(t_0+h)$, we expand ${\bf y}$ around $t_0$ and obtain
\begin{equation}
\label{eq:taylorseries}
{\bf y}(t_0 +h) = {\bf y}(t_0) + {\bf y}'(t_0) h + {1\over
{2!}} {\bf y}''(t_0) h^2 + \cdots + {1 \over {k!}}{\bf
y}^{(k)}(t_0) h^k + \cdots
\end{equation}e
A numeric approximation of ${\bf y}(t_0+h)$ is obtained by truncating
(\ref{eq:taylorseries}) at a pre-determined order.


The main problem connected with the Taylor method is the need to
compute high-order derivatives ${\bf y}'',{\bf y}''',\ldots, {\bf
  y}^{(k)}$ at $t _0$.

\subsection*{Van der Pol's Equation}
To illustrate how to derive an integration scheme using the Taylor
method, let's look at a special case of the famous Van der Pol's
equation
\begin{eqnarray}
\label{eq:vanderpol}
x'&=&y\cr
y'&=&(1-x^2)y-x
\end{eqnarray}
with initial value $(x, y) = (2, 0)$.  The second and third order
derivatives of $x, y$ with respect to time are
\begin{eqnarray}
x''&=& (1-x^2)y-x\cr
y''&=& x^3 -x  - 2 x y^2 + (x^4 - 2 x^2)y\cr
x'''&=& x^3 -x  - 2 x y^2 + (x^4 - 2 x^2)y\cr
y'''&=& 2x^3-x^5 + (-1+ 5x^2+3x^4-x^6)y +(-8x+4x^3)y^2 -2y^3
\end{eqnarray}
Hence a third order Taylor method for the initial value problem
(\ref{eq:vanderpol}) is
\begin{eqnarray*}
\left(\matrix{x_{n+1}\cr y_{n+1}}\right) &=&
\left(\matrix{x_n\cr y_n}\right) +
\left(\matrix{y_n\cr (1-x_n^2)y_n-x_n}\right) h\cr
 & & \hspace*{-16mm}
 +{1\over {2!}} \left(\matrix{(1-x_n^2)y_n-x_n\cr
   x_n^3-x_n-2x_ny_n^2+(x_n^4-2x_n^2)y_n}\right) h^2 \cr
 & & \hspace*{-16mm}
 +{1\over {3!}} \left(\matrix{x_n^3-x_n-2x_ny_n^2+(x_n^4-2x_n^2)y_n\cr
    2x_n^3-x_n^5+(-1+5x_n^2+3x_n^4-x_n^6)y_n+(-8x_n+4x_n^3)y_n^2-2y_n^3}
\right) h^3\cr
\left(\matrix{x_0\cr y_0}\right) &=&
\left(\matrix{2\cr 0}\right)
\end{eqnarray*}
As one can see from these equations, expressions for higher order
derivatives are quite complicated, and the complexity increases
dramatically as order increases.  This difficulty is precisely the
reason that Taylor method is not widely used.


Fortunately, for initial value problems where $f$ is composed of
polynomials and elementary functions, the higher order derivatives can
be generated automatically. In fact, this is precisely the motivation
of writing \taylorname{}.

\subsection*{Automatic Generation of Taylor Coefficients}

The algorithm for computing Taylor coefficients recursively has been
known since the 60s and is commonly referenced as {\it automatic
  differentiation} in the literature. It has been employed in software
packages such as ATOFMT.  A detailed description of the algorithm can
be found in \cite{JZ} (see more references therein).  Here we give a
brief account of the idea involved.


Let $f(t)$ be an analytic function and denote the $i$th Taylor
coefficient at $t_0$ by
\begin{eqnarray*}
(f)_i = {{f^{i}(t_0)} \over {i!}}
\end{eqnarray*}
where $f^{i}(t)$ is the $i$th derivative of $f$ at $t_0$. 
The Taylor expansion of $f(t)$ around $t_0$ can be
conveniently expressed as
\begin{eqnarray*}
f(t_0 +h) = (f)_0 + (f)_1 h + (f)_2 h^2 + \cdots + (f)_n h^n  \cdots
\end{eqnarray*}

Let $(p)_i, (q)_i$ be the $i$th Taylor coefficients of $p, q$ at
$t_0$. The Taylor coefficients for $p\pm q$, $pq$ and $p/q$ can be
obtained recursively using the following rules.

\begin{eqnarray}
\label{eq:recursivea}
(p\pm q)_i &=& (p)_i \pm (q)_i \cr
(p q)_i &=& \sum_{r=0}^{i} (p)_r (q)_{i-r} \cr
\left({p\over q}\right)_i &=& {1\over q}\left\{ 
(p)_i - \sum_{r=1}^{i} (q)_r \left({p\over q}\right)_{i-r}\right\} 
\end{eqnarray}
To compute the Taylor coefficients for (\ref{eq:ivp}), one first
decomposes the right hand side of the differential equation into a
series of simple expressions by introducing new variables, such that
each expression involves only one arithmetic operation. These
expressions are commonly called {\it code lists}. One then uses the
recursive relations (\ref{eq:recursivea}) and the initial values to
generate the Taylor coefficients for all the the variables.

For example, the Van der Pol equation (\ref{eq:vanderpol}) can be
decomposed as
\begin{eqnarray*}
& & u_1 = x,\;\;  u_2 = y,\;\; u_3 = 1,\;\; u_4= u_1 u_1\cr
& & u_5 = u_3-u_4,\;\; u_6 = u_5 u_2,\;\; u_7= u_6-u_1\cr
& & u_1' = u_2, \;\; u_2'= u_7
\end{eqnarray*}
Using the initial value $(x_0, y_0)=(2, 0)$, the Taylor coefficients
of all $u_i$s can be easily generated using (\ref{eq:recursivea}).


The Taylor coefficients for elementary functions can also be generated
recursively. Some of the rules are:
\begin{eqnarray*}
(p^a)_i &=& {1 \over p} \sum_{r=0}^{i-1}\left(
a - { {r(a+1)\over i}}\right)(p)_{i-r}(p^a)_r\;\;\; \mbox{
where $a$ is a real constant} 
\end{eqnarray*}
\begin{eqnarray*}
(e^p)_i &=& \sum_{r=0}^{i-1} \left(1-{r\over i}\right)
(e^p)_r (p)_{i-r}\cr
(\ln p)_i &=& {1\over {p}} \left\{ (p)_i - \sum_{r=1}^{i-1}
\left(1 - {r\over i}\right)(p)_r(\ln p)_{i-r}\right\}
\end{eqnarray*}
\begin{eqnarray*}
(\sin p)_i &=& \sum_{r=0}^{i-1}\left({{r+1}\over
i}\right)(\cos p)_{i-1-r} (p)_{r+1}\cr
(\cos p)_i &=& -\sum_{r=0}^{i-1}\left({{r+1}\over
i}\right)(\sin p)_{i-1-r} (p)_{r+1}\cr
\end{eqnarray*}
\begin{eqnarray*}
(\tan^{-1} p)_i &=& \sum_{r=0}^{i-1}\left(1 - {r\over i}\right)
\left( {1 \over {1+p^2}}\right)_r (p)_{k-r}
\end{eqnarray*}

% \section{Jet arithmetic representations}\label{app:jetarith}

\addcontentsline{toc}{section}{\refname}
\bibliographystyle{alpha}
\begin{thebibliography}{9}
\bibitem{JZ} \`A. Jorba and M. Zou. A software package for the
  numerical integration of ODE by means of high-order Taylor
  methods. {\it Exp. Math.}, 14(1):99117, 2005.
\bibitem{GJJMZ} J. Gimeno, \`A. Jorba, M. Jorba-Cusc\'o, N. Miguel,
  and M. Zou. Numerical integration of high-order variational
  equations of odes. Revisions requested, 2022.
% \bibitem{GJZ} J. Gimeno, \`A. Jorba, and M. Zou. A software package for the numerical integration of ODE by means of high-order Taylor methods using jet transport. Work in progress, 2022.
\end{thebibliography}

\end{document}
